---
title: "gt package in R"


output: 
  html_document:
    toc : true
    number_sections: true
    toc_float : true
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(comment = NA, results='hide', message = FALSE, warning = FALSE )
library(tidyverse)
library(gt)

```



```{r include=FALSE }




Titanic = read.csv('./titanic.csv')
Titanic = sample_n(Titanic, 15)  
  

```

# introduction 

gt is used to let anyone who wants to present a table to present it nicely .

it let you add title , subtitle , labels to columns , row grouping and color the table as you wish 

## the probelm 
for example if you want to present a table in R  what  you normally do is
```{r results='markup'}
   Titanic %>% 
  select(gender,age,name,sibsp,parch,ticket,fare)
    


```

but with the gt package there are much more to do .

# gt()

<p> the fist method and the main method in the package is gt() <br> 
gt let you  present a table in nicer way  </p>
we can see here how it look 
```{r results='markup'}
Titanic %>% 
  gt() 
    
  
```
we notice that there is a slight difference but we can't yet say it's **beautiful** table

# tab header

## title

more to that we can easily write a title 
```{r results='markup'}
Titanic %>% 
  gt() %>% 
  tab_header(
    title = "Titanic" )

    
  
```
we can see the title shown in the top of the table 

## subtitle

we can put below the title a subtitle to make our table more informative 

```{r results='markup'}
Titanic %>% 
  gt() %>% 
  tab_header(
    title = "Titanic ",
    subtitle =" British luxury passenger liner that sank on April 14–15, 1912"
    
  )
```
## md  

md()  let you format the text as markdown such as making a word bold or inserting break-line 

```{r results='markup'}
Titanic %>% 
  gt() %>% 
  tab_header(
    title = md("Titanic <br> <br/> "),
    subtitle =md(" **British** luxury passenger liner that sank on April 14–15, 1912")
    
  )
```
## grouping
if we care about a column and want to present it as main column in our table we can use  groupname_col 
such as in this example we care about the class of passengers and want it to look exactly the affect of the class in our table the method will group the column depending on there class

```{r results='markup' }

Titanic %>% 
  gt( groupname_col ='pclass'	 ) %>% 
  tab_header(
    title = md("Titanic"),
    subtitle =md(" British luxury passenger liner that sank on April 14–15, 1912")  
    
  )
```

## row name column
same as the column we can make the same thing to rows if we care about one column that will add important insight to our presentation for example the gender 

```{r results='markup'}

Titanic %>% 
  gt(rowname_col = 'gender' , groupname_col ='pclass'	 ) %>% 
  tab_header(
    title = md("Titanic"),
    subtitle =md(" British luxury passenger liner that sank on April 14–15, 1912")  
    
  )
```
it will show it in the begging  and will add a separation line after it 

but it doesn't fit right there is lots of spaces we can change the group column to make it inline with the rowname by introducing `tap_options()`
```{r results='markup'}
Titanic %>% 
  gt(rowname_col = 'gender' , groupname_col ='pclass'	 ) %>% 
  tab_header(
    title = md("Titanic"),
    subtitle =md(" British luxury passenger liner that sank on April 14–15, 1912")  
    
  ) %>% tab_options(row_group.as_column = TRUE)
```
now it fit without that extra space



# fmt
format will make custom formatting to a column  there are many formatting such as formatting the crunchy or date or nulls and many other 


## formating currncy

if you see the `fare` column you can notice that there are  no currency added to it we can easily added by using the fmt_currency()
```{r results='markup'}
Titanic %>% 
  gt(rowname_col = 'gender' , groupname_col ='pclass'	 ) %>% 
  tab_header(
    title = md("Titanic"),
    subtitle =md(" British luxury passenger liner that sank on April 14–15, 1912")  
    
  ) %>% tab_options(row_group.as_column = TRUE) %>% 
  fmt_currency(columns = fare , currency = 'pound' ) ## i don't know in what currency it is so i assumed it is in pound 


```
you notice that not only it added the curnccy it also change the formate to two decimal places we can change it or remove the decimal places by adding `fmt_currency(columns = fare , currency = 'pound' , decimals	= 0)` this will make the decimal places zero 

notice: the default parameter for the decimals is 2   as shown in the table

## nulls

it wont be that good to present your presentation with NAs written in your table fmt_missing() will provide a better approach and convert NA into dashs

```{r results='markup'}
Titanic %>% 
  gt(rowname_col = 'gender' , groupname_col ='pclass'	 ) %>% 
  tab_header(
    title = md("Titanic"),
    subtitle =md(" British luxury passenger liner that sank on April 14–15, 1912")  
    
  )%>% tab_options(row_group.as_column = TRUE) %>% 
  fmt_currency(columns = fare ,currency = "pound") %>% 
  fmt_missing(columns = everything()) ##everything will take all the columns

```


## date

fmt_date()  will let you format the date you have to  other type unfourtintly our data doesn't have any date data but, 

if the input date was 2000-02-29 you can choose format it to any of these 14 format 

1-iso: 2000-02-29

2-wday_month_day_year: Tuesday, February 29, 2000

3-wd_m_day_year: Tue, Feb 29, 2000

4- wday_day_month_year: Tuesday 29 February 2000

5- month_day_year: February 29, 2000

6- m_day_year: Feb 29, 2000

7- day_m_year: 29 Feb 2000

8- day_month_year: 29 February 2000

9- day_month: 29 February

10 - year: 2000

11- month: February

12 - day: 29

13 - year.mn.day: 2000/02/29

14 - y.mn.day: 0/02/29

example to write is : 

`fmt_date(data, columns, date_style = 2)`

notice that it must be in the ISO 8601 form of  ⁠YYYY-MM-DD HH:MM:SS⁠ or YYYY-MM-DD

# moving , hiding and grouping columns
there are many thing we can do in gt package when it come to columns we can hide , move or merge columns in wasy way 

## hiding  columns

using the function `cols_hide()` you can hide any column/s you want  

```{r results='markup' }
Titanic %>% 
  gt(rowname_col = 'gender' , groupname_col ='pclass'	 ) %>% 
  tab_header(
    title = md("Titanic"),
    subtitle =md(" British luxury passenger liner that sank on April 14–15, 1912")  
    
  )%>% tab_options(row_group.as_column = TRUE) %>% 
  fmt_currency(columns = fare ,currency = "pound") %>% 
  fmt_missing(columns = everything())%>%  ##everything will take all the columns 
    cols_hide(columns = c('ticket')) 
```

## moving columns 

the function `cols_move()` let you move any col by specifying after what columns you want to move to 
```{r results='markup'}
Titanic %>% 
  gt(rowname_col = 'gender' , groupname_col ='pclass'	 ) %>% 
  tab_header(
    title = md("Titanic"),
    subtitle =md(" British luxury passenger liner that sank on April 14–15, 1912")  
    
  )%>% tab_options(row_group.as_column = TRUE) %>% 
  fmt_currency(columns = fare ,currency = "pound") %>% 
  fmt_missing(columns = everything())%>%  ##everything will take all the columns 
    cols_hide(columns = c('ticket')) %>% 
    cols_move(
    columns = c('fare'),
    after = 'embarked' 
  )
```

## grouping 
merge let you choose any columns and combine them into one column

```{r  results='markup'}
Titanic %>% 
  gt(rowname_col = 'gender' , groupname_col ='pclass'	 ) %>% 
  tab_header(
    title = md("Titanic"),
    subtitle =md(" British luxury passenger liner that sank on April 14–15, 1912")  
    
  )%>% tab_options(row_group.as_column = TRUE) %>% 
  fmt_currency(columns = fare ,currency = "pound") %>% 
  fmt_missing(columns = everything())%>%  ##everything will take all the columns 
    cols_hide(columns = c('ticket')) %>% 
    cols_move(
    columns = c('fare'),
    after = 'embarked' 
  )   %>% 
  cols_merge(
    columns = c(sibsp, parch),
    pattern = "{1}<br>{2}"
  ) 
  
```
## renaming columns 
we can see that they are combined into one column but now it's little misleading  since they under the name `sibsp` , cols_label will handle this by renaming any column we want 
```{r  results='markup'}
Titanic %>% 
  gt(rowname_col = 'gender' , groupname_col ='pclass'	 ) %>% 
  tab_header(
    title = md("Titanic"),
    subtitle =md(" British luxury passenger liner that sank on April 14–15, 1912")  
    
  )%>% tab_options(row_group.as_column = TRUE) %>% 
  fmt_currency(columns = fare ,currency = "pound") %>% 
  fmt_missing(columns = everything())%>%  ##everything will take all the columns 
    cols_hide(columns = c('ticket')) %>% 
    cols_move(
    columns = c('fare'),
    after = 'embarked' 
  )   %>% 
  cols_merge(
    columns = c(sibsp, parch),
    pattern = "{1}<br>{2}"
  )  %>% 
   cols_label( 
     sibsp = 'relatives'
     )
```


## labling columns
if you think the fist method is confusing then using tap_spanner is another way 
to create a label and map it to columns to make the table more informative 
so in our table there are 

```{r results='markup'}
Titanic %>% 
  gt(rowname_col = 'gender' , groupname_col ='pclass'	 ) %>% 
  tab_header(
    title = md("Titanic"),
    subtitle =md(" British luxury passenger liner that sank on April 14–15, 1912")  
    
  )%>% tab_options(row_group.as_column = TRUE) %>% 
  fmt_currency(columns = fare ,currency = "pound") %>% 
  fmt_missing(columns = everything())%>%  ##everything will take all the columns 
    cols_hide(columns = c('ticket')) %>% 
    cols_move(
    columns = c('fare'),
    after = 'embarked' 
  )   %>% 
  tab_spanner(
    label = "relatives",
    columns = c(sibsp	,parch) )%>% tab_options(row_group.as_column = TRUE) # parch are  number of parents / children aboard the Titanic
```
# coloring 
we can color our dataset to make it more informative 

 we can color any numerical  columns for example if we want color the fare depending about how cheap 
```{r  results='markup' }
Titanic %>% 
  gt(rowname_col = 'gender' , groupname_col ='pclass'	 ) %>% 
  tab_header(
    title = md("Titanic"),
    subtitle =md(" British luxury passenger liner that sank on April 14–15, 1912")  
    
  )%>% tab_options(row_group.as_column = TRUE) %>% 
  fmt_currency(columns = fare ,currency = "pound") %>% 
  fmt_missing(columns = everything())%>%  ##everything will take all the columns 
    cols_hide(columns = c('ticket')) %>% 
    cols_move(
    columns = c('fare'),
    after = 'embarked' 
  )   %>% 
  tab_spanner(
    label = "relatives",
    columns = c(sibsp	,parch) )%>% tab_options(row_group.as_column = TRUE) %>%  # parch are  number of parents / children aboard the Titanic
    data_color(
    columns = fare,
    colors = scales::col_numeric(
      palette = c("green", "red"),
      domain = c(7, 250)
    ))

```
# sourse and footnote 
with gt you can add the source and footnote in the bottom of the table 


## sourse

```{r results='markup'}
Titanic %>% 
  gt(rowname_col = 'gender' , groupname_col ='pclass'	 ) %>% 
  tab_header(
    title = md("Titanic"),
    subtitle =md(" British luxury passenger liner that sank on April 14–15, 1912")  
    
  )%>% tab_options(row_group.as_column = TRUE) %>% 
  fmt_currency(columns = fare ,currency = "pound") %>% 
  fmt_missing(columns = everything())%>%  ##everything will take all the columns 
    cols_hide(columns = c('ticket')) %>% 
    cols_move(
    columns = c('fare'),
    after = 'embarked' 
  )   %>% 
  tab_spanner(
    label = "relatives",
    columns = c(sibsp	,parch) )%>% tab_options(row_group.as_column = TRUE) %>%  # parch are  number of parents / children aboard the Titanic
    data_color(
    columns = fare,
    colors = scales::col_numeric(
      palette = c("green", "red"),
      domain = c(7, 250)
    ))  %>%  
  tab_source_note(
    source_note = md(
      " sourses  and for more inforomation about gt : [youtube video](https://www.youtube.com/watch?v=aPRVzk9pvno&t=1650s) , [article](https://gt.rstudio.com/articles/case-study-gtcars.html#hiding-and-moving-some-columns) and [article2 ](https://gt.rstudio.com/articles/case-study-gtcars.html)")
  ) #with help of md method
```
## footnote

adding a footnote is nice way to write more information about a specific column 

```{r results='markup' }
Titanic %>% 
  gt(rowname_col = 'gender' , groupname_col ='pclass'	 ) %>% 
  tab_header(
    title = md("Titanic"),
    subtitle =md(" British luxury passenger liner that sank on April 14–15, 1912")  
    
  )%>% tab_options(row_group.as_column = TRUE) %>% 
  fmt_currency(columns = fare ,currency = "pound") %>% 
  fmt_missing(columns = everything())%>%  ##everything will take all the columns 
    cols_hide(columns = c('ticket')) %>% 
    cols_move(
    columns = c('fare'),
    after = 'embarked' 
  )   %>% 
  tab_spanner(
    label = "relatives",
    columns = c(sibsp	,parch) )%>% tab_options(row_group.as_column = TRUE) %>%  # parch are  number of parents / children aboard the Titanic
    data_color(
    columns = fare,
    colors = scales::col_numeric(
      palette = c("green", "red"),
      domain = c(0, 250)
    ))  %>%  tab_source_note(
    source_note = md(
      " sourses  and for more inforomation about gt : [youtube video](https://www.youtube.com/watch?v=aPRVzk9pvno&t=1650s) , [article](https://gt.rstudio.com/articles/case-study-gtcars.html#hiding-and-moving-some-columns) and [article2 ](https://gt.rstudio.com/articles/case-study-gtcars.html)")
  ) %>% 
    tab_footnote(
    footnote = " parch are  number of parents / children aboard the Titanic.",
    locations = cells_column_labels(columns = parch) ## the gt package  cells_column_labels is a Location helper for targeting the column labels
  )

```

# END
in the end there more to do with gt such as adjusting width of the column , changing the font of the table content or ordering the rows and changing the color of the background with a very small steps , above i provided links if you want to go deeper 

**thank you** for your listening 